---
title: 'Assignment 5'
author: "Guy Montag"
subtitle: Basic Skills in Statistics | 2022
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

*** 

**Topics: Probabilities \& Distributions \& Central Limit Theorem**

In the course syllabus you can find the deadline for the assignment and the suggested literature.

*** 

### Probabilities

##### 1. 

*What is meant with probabilities being `mutually exclusive'?*

Mutual exclusivity means that events cannot happen at the same time but are either/or events. 

##### 2. 

*Think of a (fictional) example for which the following statement holds: $P(A|B) = P(A)$.*

Throwing a die and flipping a coin. (elaborate more)

P(Heads or Tails) = 1
P(Rolling any number of a die) = 1

P(Heads or Tails|Rolling any number of a die) = 1 == P(Heads or Tails) = 1

The events are independent of each other

##### 3. 

*Show, using a (fictional) example, that the following statement holds: $P(A|B) \neq P(B|A)$.*

The probability of the stock market crashing given that Russia invades Ukraine is unequal to the probability of the Russia invading Ukraine given that the Stock market is crashing. 
This can be proven by looking at the past history. In March 2020, 2007/2008 the stock market crashed but Russia didn't invade Ukraine. However given the invasion of ukraine four days 
ago the stock market crashed as well. Therefore $P(A|B) \neq P(B|A)$ holds. 

##### 4.

*Calculate the probability of selecting a PM (Psychological Methods) student that passed mathematics in high school from the total population of 3rd year Psychology students ($P(passedMath \; and \; PM)$). Consider the following facts: population = 400; # students that passed high school math = 80; # PM students = 20; # PM students that passed math = 15.*

15/400 = 0.0375

##### 5.

*Simulate the selection of 1 student from the population of 400 3rd year Psychology students ($(N_{passedMath} = 80; N_{PM} = 20)$), and do this a 1000 times. To do this, download* \texttt{assignment1.Rdata} *(from Canvas) and load it to your workspace in R (use the * \texttt{load()} * function and check with* \texttt{ls()} *). How many PM students, students that passed math, and PM students that passed math did you select? The following script gives you the right direction. Does the result confirm your expectations?*

```{r, eval=TRUE}
load("assignment1.Rdata")

n_PM <- n_m <- n_PM_m <- 0
for(i in 1:1000) {
  s <- sample(1:nrow(studentDat),1,replace = T)
  if(studentDat[s,]$PM == 'yesPM' & studentDat[s,]$passedMath == 'passedMath') {
  n_PM_m <- n_PM_m + 1
  }
  if(studentDat[s,]$PM == 'yesPM') {
    n_PM <- n_PM + 1
  }
  if(studentDat[s,]$passedMath == 'passedMath') {
    n_m <- n_m + 1
  }
}

n_m
n_PM
n_PM_m
n_PM_m/1000
```

The result does confirm the expectations as around 3-4 percent of the students are PM Students and passed the Math test.

##### 6.

*What is the probability of selecting two PM students in a sample of 2 (without replacement)? Are the probabilities of the first and second selection independent? Explain.*

(15/400) * (14/399) = 0.001315789 

The probilities are not independent because one student gets permanently removed. 

##### 7.

*Check your calculation using a simulation. Give the R code and explain it.*

```{r, eval=T}

sampling_students <- function(x) {
  m_students <- matrix(replicate(x,sample(studentDat$PM,2,replace = F)),ncol = 2,byrow = T)
  counter_PM <- 0
  for(i in 1:x) {
    if(m_students[i,1] == "yesPM" & m_students[i,2] == "yesPM" ) {
      # print(i)
      counter_PM <- counter_PM + 1
  }
  }
  return(counter_PM/x)
}

sampling_students(1000)
```
Code Explanation: 

The function sampling_students calculates the probability that two sequentially selected students are both in the Psychological Methods Program. To do so it samples 2 students without replacement from the dataset [studentDat$PM] 1000 (variable parameter) times. The draws sample will be in the from of a data frame with 1 row and 2 columns. Subsequently, it will assess, 
whether both values of the data frame are indeed PM students. If that is the case it will increase the counter by 1. Ultimately it returns the proportion of samples from 1000 (variable parameter) that have fulfilled the condition. 

The result is congruent with the calculated probability in question 6. 

*** 

### Probability Distributions

The following questions are based on the students taking an Xhosa exam. The exam consist of 10 two-choice questions. None of the students understands a single Xhosa word. Using R we can simulate scores of 250 students as follows. And using some of our plotting skills:
```{r}
scores <- rbinom(250, 10, .5)
plot(table(scores))
```

##### 8.

*What is the (theoretical) probability that a student answers a random question correctly? Why?*

The prob is 1/2. This is the case because the test has only two answers per question. Therefore it behaves like a coin flip for answering a random question
correctly. This only holds under the assumption that there is no way the student can obtain any knowledge on the question but is purely guessing the answer.

##### 9.

*Is the (theoretical) probability of obtaining the series 0101010101 smaller, bigger, or equal to obtaining the series 1111111111? Why?*

It is the same probability, if the events/questions are independent of each other. In that case the questions behave like 10 consecutive coin flips. The likelihood 
of obtaining the series 0101010101 (in coin terms T(ail)H(ead)THTHTHTH) is therefore the same as 1111111111 (HHHHHHHHHH). 

##### 10.

*Is the (theoretical) probability of obtaining sum score 5 smaller, bigger, or equal to obtaining sum score 10? Why?*
```{R}
p = 0.5
prob_10_question <- p^10

# therefore: 

prob_10_out_of_10 <- prob_10_question * choose(10,10)

prob_5_out_of_10 <- prob_10_question * choose(10,5)

print(prob_10_out_of_10)
print(prob_5_out_of_10)
```
They are are not equal.

Theoretical answer: 

The theoretical probability of obtaining a sum score of 5 is larger than the theoretical probability of obtaining a sum score of 10.
While there is only one possible way of obtaining a score of 10 (choose(10,10)), there are 252 (choose(10,5)) possibilities to get are sum score of 5.

##### 11.

*How many different series have sum score 6? Calculate this using a quick function in R.*
```{R}
choose(10,6)
```
There are 210 possibilities to get a score of 6.

##### 12.

*What is the (theoretical) probability of sum score 5? Calculate this using the binomial formula (and the choose function), and using the dbinom function in R.*
```{R}
# dbinom
dbinom(5, 10, 0.5)

# binominal formula 

choose(10,5)*0.5^5*0.5^5
```

Both solutions yield the same answer. 

##### 13.

*It's found that the meaning of a few words in Xhosa can be deduced from English. The probability of answering the two-choice questions correctly is now $p\left(correct\right)=.75$. You don't need to simulate this using the quincunx.*

*What is the (theoretical) probability of sum score 5? Why is it lower, higher, or equal to Q11*
```{R}
dbinom(5,10,0.75)
```
The likelihood is lower, because now it is easier to get more scores right. 

##### 14.

*What's the (theoretical) probability of sum score 9 or higher? Calculate this using a formula (use R as a calculator), and show a quick function in R.*
```{R}
# easy R way
sum(dbinom(9:10,10,0.75))

# Manually calculated

a <- choose(10,9) * (0.75^9) * (0.25^(1))

b <- choose(10,10) *(0.75^10)

sum(a,b)
```

Both ways yield the same result. qed.

*** 

\newpage


### Central limit theorem (CLT)

Watch the graph below from http://www.washingtonpost.com/blogs/wonkblog/wp/2014/09/25/think-you-drink-a-lot-this-chart-will-tell-you/. 

```{r, out.width = "90%", echo = FALSE}
library(png)      # For grabbing the dimensions of png files
library(knitr)    # For knitting document and include_graphics function
img1_path <- "fig-nyt.png"
img1 <- readPNG(img1_path, native = TRUE, info = TRUE)
include_graphics(img1_path)
```

##### 15.

*What distribution provides an accurate description of the means of the number of alcoholic beverages that an American drinks, when we ask 100 random Americans? Specifically, what is the sampling distribution of the mean of 100 observations (each of these 100 observations being the average alcoholic intake of a single person). So the average of the average intake, across the sample of 100 Americans. The distribution of the average intake per person is the data distribution (shown in the image), the distribution of the average intake for a sample of 100 persons is the sampling distribution. This questions asks you what the sampling distribution looks like.*

The sampling distribution of the sample means is always a normal distribution around the value of the mean of the population distribution.

##### 16.

*Explain what **population distributions**, **sample distributions**, and **distributions of sample means** are? Clearly describe how these differ.*

The population distribution is the distribution that would be constructed of all the elements of the observation. If I would try to make a claim about students in the Netherlands, the population distribution would be the distribution of all students in the Netherlands. This distribution doesn't change as it includes all elements possible. The sampling distribution is the distribution of the subset of the population that were taken from the population. It changes from sample to sample. The distributions of sample means would be the distribution of the means (statistic) if not one sample but more that one sample would have been taken. In case of 1000 samples there would be 1000 sample means. The distributions of sample means would be the distribution of these sample means. Due to the Central Limit Theorem, the sampling distribution of the sample means is normally distributed around the population mean regardless of the shape of the population distribution. The variance and standard error depend on the sample size of the samples taken. 

##### 17.

*Read up to chapter 3.1. Why do you expect that the sample mean is close to the population mean?*

Following the central Limit Theorem, the mean of the sample will be closer to the mean of the population than any other number. This is because the distribution of the means is normally distributed around the population mean. The normal distribution implies that a sample statistic that is close to the mean has a higher probability than one that is far away. Therefore there is a natural tendency towards the sample mean, which is the closest approximation of the population mean. Therefore it is to be expected that the sample mean is close to the population mean. 

##### 18.

*Plot a distribution of sample means (choose your favorite distribution). Show what happens to this distribution if you increase or decrease the sample sizes of those samples.*
```{R}
multiplot <- function(row, col) {
  par(mfrow = c(row, col), pty = "s")
}

sampling_distribution <- function(x,y,z) { # sample from a noraml distribution. 
  means <- numeric()
  for(i in 1:x) {
    s <- sample(1:y,z,replace = T)
    m <- mean(s)
    means <- c(means,m)
  }
  return(means)
}


sampling_distribution_2 <- function(x,y,z) { # show that the CLT holds for other distribution types.
  means_2 <- numeric()
  for(i in 1:x) {
    s <- rexp(y) # exponential distribution
    m <- mean(s)
    means_2 <- c(means_2,m)
  }
  return(means_2)
}

multiplot <- function(row, col) {
  par(mfrow = c(row, col), pty = "s")
}

multiplot(2,2)


x <- sampling_distribution(1000,100,10)
y <- sampling_distribution(1000,100,1000)

x_2 <- sampling_distribution_2(1000,10)
y_2 <- sampling_distribution_2(1000,1000)



hist(x,main = "Hist of mean from sample() w/ n = 10")
hist(y,main = "Hist of mean from sample() w/ n = 1000")

hist(x_2,main = "Hist of mean from exp() w/ n = 10")
hist(y_2,main = "Hist of mean from exp() w/ n = 1000")
```

It become apparent form the two histograms above that the distribution is still normal but that larger sample sizes (n) result in a much narrower variance and standard error. 


##### 19.

*Show, based on the formula of the standard error, what happens with the SE if you increase the sample size.*

The standard error decreases with increasing n: 

$SE = \frac{sd}{\sqrt{n}}$ so if n increases the root gets larger, therefore the denominator gets larger, therefore the fraction - the SE - gets smaller. This will in effect make the distribution narrower. 

##### 20.

*Why is the SE dependent on the sample size? Clearly explain. Hint: compare the distribution of the sample means with the population distribution.*

The SE is dependent on the sample size because the SE is inversely related to the Standard Error. With a higher sample size the mean of each of the samples holds more precise information about the population mean. Higher precision is equal to a lower error in the sample. The sampling distribution of the sample means implies that a sample statistic that is close to the mean has a higher probability than one that is far away. With higher sample size the sampling distribution of the sample means becomes narrower. Therefore the confidence in the statistic becomes higher and the error becomes lower. 

*** 

### Vasishth \& Broe

A **selection of** Vasishth \& Broe's questions about the literature.

##### 1.

*Imagine that you have a biased coin, where the probability of obtaining a heads is not 0.5 but 0.1. When the coin is tossed four times, what are the probabilities of obtaining 0, 1, 2, 3, 4 heads? [Tip: Use R to do the calculations for you!]*
```{R}
dbinom(0:4,4,0.1)
```

The probability for obtaining 0 is 0.6561, 1 is 0.2916, 2 is 0.0486, 3 is 0.0036, 4 is 0.0001. 

##### 2.

*What is the probability of obtaining any of the numbers 2, 4, or 6 if a die is tossed three times in a row?*
```{R}
p_246 <- 1/6+1/6+1/6

1 - dbinom(3,3,0.5)
```
